---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = F,
  message = F,
  fig.align = "center"
)
```

## Installation

Install the package from [r-universe](http://tbep-tech.r-universe.dev/ui/#builds) as follows. The source code is available on [GitHub](https://github.com/tbep-tech/tbeploads).

```{r eval = F}
# Install tbeploads in R:
install.packages('tbeploads', repos = c('https://tbep-tech.r-universe.dev', 'https://cloud.r-project.org'))
```

Load the package in an R session after installation:

```{r}
library(tbeploads)
```

## Usage

Load estimates are broadly defined as domestic point source (DPS), industrial point source (IPS), material losses (ML), nonpoint source (NPS), atmospheric deposition (AD), and groundwater sources and springs (GW).  The functions are built around these sources with unique inputs for each.  See the articles below for detailed documentation on each loading type.

| Loading type | Article |
|---|---|
| Domestic point source | [DPS](dps.html) |
| Industrial point source | [IPS](ips.html) |
| Material losses | [ML](ml.html) |
| Atmospheric deposition | [AD](ad.html) |
| Non-point source | [NPS](nps.html) |
| Springs | [SPR](spr.html) |
| Groundwater | [GW](gw.html) |
