---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = F, 
  message = F, 
  fig.align = "center"
)
```

## Installation

Install the package from [r-universe](http://tbep-tech.r-universe.dev/ui/#builds) as follows. The source code is available on [GitHub](https://github.com/tbep-tech/tbeploads).

```{r eval = F}
# Install tbeploads in R:
install.packages('tbeploads', repos = c('https://tbep-tech.r-universe.dev', 'https://cloud.r-project.org'))
```

Load the package in an R session after installation:

```{r}
library(tbeploads)
```

## Usage

Load estimates are broadly defined as domestic point source (DPS), industrial point source (IPS), nonpoint source (NPS), atmospheric deposition (AD), and groundwater sources and springs (GW).  The functions are build around these source with unique inputs for each. 

### DPS

The DPS functions are designed to work with raw entity data provided by partners.  The core function is `anlz_dps_entity()` that requires only a vector of file paths as input, where each path points to a file with monthly parameter concentration and and flow data.  The data also describe whether the observations are for end of pipe (direct inflow to the bay) or as reuse (applied to the land).  Different calculations are used to determine the load for each.  The file names must follow a specific convention, where metadata for each entity is found in the `facilities` file included with the package using information in the file name. 

For convenience, three example files are included with the package.  These files are used as input to the function.  Non-trivial data preprocessing is needed for each file and those included in the package are the correct format. 

```{r}
fls <- list.files(system.file('extdata/', package = 'tbeploads'),
  pattern = '\\.txt$', full.names = TRUE)
anlz-dps_entity(fls)
```

